Generalized Bradley-Terry models

### 2025-03-21 inst/jss_hyper3_real.Rnw
jss house style whitespace

\subsection{The package in use}
Package idiom for creating a likelihood function for this dataset is
straightforward.  Considering first only Rangers-Celtic matches:
%
<<defineoldfirm, cache=TRUE>>=
oldfirm <- function(lambda){
    H <- hyper3(pnames = c("Rangers", "Celtic"))
    H[c(Rangers = lambda            )] %<>%      add(81     )
    H[c(                  Celtic = 1)] %<>%      add(     42)
    H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache=TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
lam <- seq(from=0.9, to=3, len=howmany)
logLike <- sapply(lam,f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

\begin{figure}[htbp]
  \begin{center}
<<plotproflikeoldfirm, echo=FALSE>>=
plot(lam,logLike - max(logLike),type='b')
abline(v=1,col='gray')
segments(x0=1.785,y0=0.1,y1=-1,col='red')
abline(h = c(0,-2),col='gray')
@
\caption{A\label{profilesupportlambda} profile support function for
  the home-team advantage $\lambda$.  Vertical gray line shows
  $H_0\colon\lambda =1$, horizontal lines show the
  two-units-of-support limits; small red bar shows
  $\hat{\lambda}\simeq 1.785$}
\end{center}
\end{figure}

<<optimoldfirm, cache=TRUE>>=
(lambda_max <-  optimize(f,c(1.5,2.0),maximum=TRUE)$maximum)
@

Then using this value we may estimate the Bradley-Terry strengths of
Rangers and Celtic:
%
<<mleoldfirm,cache=TRUE>>=
maxp(oldfirm(lambda_max))
@
%
Compare with the theoretical value of $\widehat{p_{\mbox{\tiny
      C}}}=\frac{\sqrt{42\cdot 76}}{\sqrt{81\cdot 46}+\sqrt{42\cdot
    76}}\simeq 0.4807$, $\widehat{p_{\mbox{\tiny R}}}
=\frac{\sqrt{81\cdot 46}}{\sqrt{81\cdot 46}+\sqrt{42\cdot 76}}\simeq
0.5193$, and $\widehat{\lambda}=\frac{\sqrt{76\cdot 81}}{\sqrt{46\cdot
    42}}\simeq 1.785$.

@@

\begin{table}[t!]
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}
%
We are now in a position to consider the entirety of
Table~\ref{oldfirm}.  Bespoke package function
\fct{home\_draw\_away3} converts data in the form of
Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
Table~\ref{windrawloss}; it takes numerical values for the home
advantage $\lambda$ and proclivity to draw $D$:
%
<<eg_RCLF3_lf, cache=TRUE>>=
home_draw_away3(RCLF3_table, lambda=1.7, D=0.3)
@
%
Observe that the resulting likelihood function has two distinct types
of parameters: the plain Bradley-Terry strengths $\mathbf{p}=(p_R,
p_C,p_L,p_F)$, with unit sum; but also the auxiliary parameters
$\lambda,D$.  To make inferences about these parameters we calculate a
profile likelihood function: conditional on $\lambda,D$ we maximize
the likelihood over possible values of $\mathbf{p}$.
%
<<defineF>>=
f <- function(v){
    H <- home_draw_away3(RCLF3_table, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
@
%
Thus \fct{f} gives us a likelihood function for
$\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
\fct{optim}:
%
<<findmaxlambdaD, cache=TRUE>>=
maxv <- optim(par=c(2, 0.3), fn=f, control=list(fnscale = -1))$par
maxv
@
%
Then
%
<<showevaluate, cache=TRUE>>=
maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D=maxv[2]))
@
%
Such estimates have rich predictive power.  For example, if Rangers
were to play Celtic with both teams away [thus nullifying the home
  ground advantage: $\lambda=1$ in Table~\ref{windrawloss}], the
probability of win/draw/loss for Rangers would be
$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_C}{(1+D)(p_R+p_C)}\right)$,
estimated at $(0.39;0.27;0.34)$.
%
<<makecontourdata, echo=FALSE, cache=TRUE>>=
n <- howmany
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
@

\begin{figure}[htbp]
  \begin{center}
<<docontourplot, cache=FALSE>>=
contour(lambda, D, matrix(LL, n, n), levels = -2*(1:9),
        xlab = expression(lambda), ylab = expression(D))
points(maxv[1], maxv[2], pch = 16, col = 'red')
abline(v=1, col = 'gray')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate.  Vertical gray line shows $H_0\colon\lambda=1$}
  \end{center}
\end{figure}

<<testlambdazero, cache=TRUE, include=FALSE>>=
jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \fct{RB\_BB\_LF} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \fct{RB\_BB\_LF} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \fct{oldfirm} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@ 
%
Above, we define function \fct{o} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \fct{o} as a
support function for $\lambda$, and then maximize it:
%
<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$~value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo=FALSE>>=
if(FALSE){
    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L-max(L), type="b")
    abline(h=c(0, -2))
    abline(v=0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e=sushi_eq_classes, h=c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi,echo=FALSE, cache=TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H,give=1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s,f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi,echo=FALSE, cache=TRUE>>=
osushi <- optimize(f, c(2,4), maximum=TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo=FALSE, cache=TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo=FALSE, cache=TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0,1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1,2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$~value of about
$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.

\begin{figure}[htbp]
  \begin{center}
<<plotlikesushi, echo=FALSE>>=
plot(log(s),L-max(L),type='b',
     xlab=expression(log(lambda)), ylab='Log-likelihood')
abline(v=0,col='gray')
abline(h=c(0,-2),col='gray')
segments(x0 = log(sushimaxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(sushi_upper    ), y0 = -2.2, y1 = -1.8, col = 'blue')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$ in the case of sushi preferences.
  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
  show the two-units-of-support limits with the credible interval of
  $(\Sexpr{round(sushi_lower,2)},\Sexpr{round(sushi_upper,2)})$,
  marked in blue.  Red bar shows evaluate at
  $\lambda\simeq\Sexpr{round(sushimaxlambda,2)}$}
\end{center}
\end{figure}

@@

Typically when a player is killed, the team suffers a number of
strategic disadvantages; player deaths are thus a quantifiable proxy
for both individual ability and also both teams' tactical execution.
Here I analyse the order statistic for number of deaths of each of the
five {\em Cloud9} players.
%
<<echo=FALSE>>=
a <- read.table("cloud9_boston_2018.txt", header=TRUE)
a[,-1]
@
%
From the first line we see that {\em Cloud9} played {\em FaZe Clan} on
the {\em inferno} map, won the match, and \skad\ had the least number
of deaths, followed by \rush, then \tarik, and so on.  Consider two
{\em Cloud9} players: \skad\ and \stew.  These teammates, although
comparable in skill [we fail to reject $H_0\colon p_{\mbox{\tiny\sc
      Skadoodle}} = p_{\mbox{\tiny\sc Stewie2k}}$, Plackett-Luce,
  $p=0.28$] have playing styles that stand in dramatic contrast.
Commentators characterise \skad\ as emphasising reactive map control
while \stew\ is more of a disruptive presence engaging in
high-variance play.

>>>>

\subsection{The package in use}
Package idiom for creating a likelihood function for this dataset is
straightforward.  Considering first only Rangers-Celtic matches:
%
<<defineoldfirm, cache = TRUE>>=
oldfirm <- function(lambda){
    H <- hyper3(pnames = c("Rangers", "Celtic"))
    H[c(Rangers = lambda            )] %<>%      add(81     )
    H[c(                  Celtic = 1)] %<>%      add(     42)
    H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache=TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda), give = 1)$`Log-likelihood`}
lam <- seq(from = 0.9, to = 3, len = howmany)
logLike <- sapply(lam, f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

\begin{figure}[htbp]
  \begin{center}
<<plotproflikeoldfirm, echo=FALSE>>=
plot(lam, logLike - max(logLike), type = 'b')
abline(v = 1, col = 'gray')
segments(x0 = 1.785, y0 = 0.1, y1 = -1, col = 'red')
abline(h = c(0, -2), col = 'gray')
@
\caption{A\label{profilesupportlambda} profile support function for
  the home-team advantage $\lambda$.  Vertical gray line shows
  $H_0\colon\lambda =1$, horizontal lines show the
  two-units-of-support limits; small red bar shows
  $\hat{\lambda}\simeq 1.785$}
\end{center}
\end{figure}

<<optimoldfirm, cache=TRUE>>=
(lambda_max <-  optimize(f, c(1.5, 2.0), maximum = TRUE)$maximum)
@

Then using this value we may estimate the Bradley-Terry strengths of
Rangers and Celtic:
%
<<mleoldfirm, cache=TRUE>>=
maxp(oldfirm(lambda_max))
@
%
Compare with the theoretical value of $\widehat{p_{\mbox{\tiny
      C}}}=\frac{\sqrt{42\cdot 76}}{\sqrt{81\cdot 46}+\sqrt{42\cdot
    76}}\simeq 0.4807$, $\widehat{p_{\mbox{\tiny R}}}
=\frac{\sqrt{81\cdot 46}}{\sqrt{81\cdot 46}+\sqrt{42\cdot 76}}\simeq
0.5193$, and $\widehat{\lambda}=\frac{\sqrt{76\cdot 81}}{\sqrt{46\cdot
    42}}\simeq 1.785$.

@@

\begin{table}[t!]
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}
%
We are now in a position to consider the entirety of
Table~\ref{oldfirm}.  Bespoke package function
\fct{home\_draw\_away3} converts data in the form of
Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
Table~\ref{windrawloss}; it takes numerical values for the home
advantage $\lambda$ and proclivity to draw $D$:
%
<<eg_RCLF3_lf, cache = TRUE>>=
home_draw_away3(RCLF3_table, lambda = 1.7, D = 0.3)
@
%
Observe that the resulting likelihood function has two distinct types
of parameters: the plain Bradley-Terry strengths $\mathbf{p}=(p_R,
p_C,p_L,p_F)$, with unit sum; but also the auxiliary parameters
$\lambda,D$.  To make inferences about these parameters we calculate a
profile likelihood function: conditional on $\lambda,D$ we maximize
the likelihood over possible values of $\mathbf{p}$.
%
<<defineF>>=
f <- function(v){
    H <- home_draw_away3(RCLF3_table, lambda = v[1], D = v[2])
    maxp(H, give = 1)$`Log-likelihood`
}
@
%
Thus \fct{f} gives us a likelihood function for
$\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
\fct{optim}:
%
<<findmaxlambdaD, cache = TRUE>>=
maxv <- optim(par=c(2, 0.3), fn = f, control = list(fnscale = -1))$par
maxv
@
%
Then
%
<<showevaluate, cache = TRUE>>=
maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D = maxv[2]))
@
%
Such estimates have rich predictive power.  For example, if Rangers
were to play Celtic with both teams away [thus nullifying the home
  ground advantage: $\lambda=1$ in Table~\ref{windrawloss}], the
probability of win/draw/loss for Rangers would be
$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_C}{(1+D)(p_R+p_C)}\right)$,
estimated at $(0.39;0.27;0.34)$.
%
<<makecontourdata, echo = FALSE, cache = TRUE>>=
n <- howmany
lambda <- seq(from = 1, to = 3, len = n)
D <- seq(from = 0.1, to = 0.6, len = n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
@

\begin{figure}[htbp]
  \begin{center}
<<docontourplot, cache = FALSE>>=
contour(lambda, D, matrix(LL, n, n), levels = -2*(1:9),
        xlab = expression(lambda), ylab = expression(D))
points(maxv[1], maxv[2], pch = 16, col = 'red')
abline(v=1, col = 'gray')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate.  Vertical gray line shows $H_0\colon\lambda=1$}
  \end{center}
\end{figure}

<<testlambdazero, cache = TRUE, include = FALSE>>=
jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \fct{RB\_BB\_LF} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C = 1, T = 2, RB = 3, BB = 3, W = 4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \fct{RB\_BB\_LF} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \fct{oldfirm} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give = 1)$`Log-likelihood`} 
@ 
%
Above, we define function \fct{o} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \fct{o} as a
support function for $\lambda$, and then maximize it:
%
<<label = maxlikesim, cache = TRUE>>=
(osup <- optimize(o, c(10, 40), maximum = TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label = LRT_RB, cache = TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$~value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo = FALSE>>=
if(FALSE){
    lambda <- exp(seq(from = log(0.9), to = log(67), len = 19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L - max(L), type = "b")
    abline(h = c(0, -2))
    abline(v = 0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e = sushi_eq_classes, h = c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi, echo = FALSE, cache = TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H, give = 1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s, f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi, echo = FALSE, cache = TRUE>>=
osushi <- optimize(f, c(2, 4), maximum = TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo = FALSE, cache = TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo = FALSE, cache = TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0, 1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1, 2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$~value of about
$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.

\begin{figure}[htbp]
  \begin{center}
<<plotlikesushi, echo = FALSE>>=
plot(log(s), L - max(L), type = 'b',
     xlab = expression(log(lambda)), ylab = 'Log-likelihood')
abline(v = 0, col = 'gray')
abline(h = c(0, -2), col = 'gray')
segments(x0 = log(sushimaxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(sushi_upper    ), y0 = -2.2, y1 = -1.8, col = 'blue')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$ in the case of sushi preferences.
  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
  show the two-units-of-support limits with the credible interval of
  $(\Sexpr{round(sushi_lower,2)},\Sexpr{round(sushi_upper,2)})$,
  marked in blue.  Red bar shows evaluate at
  $\lambda\simeq\Sexpr{round(sushimaxlambda,2)}$}
\end{center}
\end{figure}

@@

Typically when a player is killed, the team suffers a number of
strategic disadvantages; player deaths are thus a quantifiable proxy
for both individual ability and also both teams' tactical execution.
Here I analyse the order statistic for number of deaths of each of the
five {\em Cloud9} players.
%
<<echo = FALSE>>=
a <- read.table("cloud9_boston_2018.txt", header = TRUE)
a[, -1]
@
%
From the first line we see that {\em Cloud9} played {\em FaZe Clan} on
the {\em inferno} map, won the match, and \skad\ had the least number
of deaths, followed by \rush, then \tarik, and so on.  Consider two
{\em Cloud9} players: \skad\ and \stew.  These teammates, although
comparable in skill [we fail to reject $H_0\colon p_{\mbox{\tiny\sc
      Skadoodle}} = p_{\mbox{\tiny\sc Stewie2k}}$, Plackett-Luce,
  $p=0.28$] have playing styles that stand in dramatic contrast.
Commentators characterise \skad\ as emphasising reactive map control
while \stew\ is more of a disruptive presence engaging in
high-variance play.
